---
title: "Mushroom Dataset Analysis"
author: "Nazima Khan"
date: "July 21, 2017"
output: html_document
---
### The mushroom classification problem is to determine whether amushroom is edible or poisonous based on its observable features . 

### Objective
* What types of machine learning models perform best on this dataset?
* Which features are most indicative of a poisonous mushroom? 

### Dataset Information:
####This data set includes descriptions of hypothetical samples corresponding to 23 species of gilled mushrooms in the Agaricus and Lepiota Family (pp. 500-525). Each species is identified as definitely edible, definitely poisonous, or of unknown edibility and not recommended. This latter class was combined with the poisonous one. The Guide clearly states that there is no simple rule for determining the edibility of a mushroom; no rule like "leaflets three, let it be" for Poisonous Oak and Ivy.

### Attribute Information:
* cap-shape: bell=b,conical=c,convex=x,flat=f, knobbed=k,sunken=s
* cap-surface: fibrous=f,grooves=g,scaly=y,smooth=s
* cap-color: brown=n,buff=b,cinnamon=c,gray=g,green=r, pink=p,purple=u,red=e,white=w,yellow=y
* bruises?: bruises=t,no=f
* odor: almond=a,anise=l,creosote=c,fishy=y,foul=f, musty=m,none=n,pungent=p,spicy=s
* gill-attachment: attached=a,descending=d,free=f,notched=n
* gill-spacing: close=c,crowded=w,distant=d
* gill-size: broad=b,narrow=n
* gill-color: black=k,brown=n,buff=b,chocolate=h,gray=g, green=r,orange=o,pink=p,purple=u,red=e, white=w,yellow=y
* stalk-shape: enlarging=e,tapering=t
* stalk-root: bulbous=b,club=c,cup=u,equal=e, rhizomorphs=z,rooted=r,missing=?
* stalk-surface-above-ring: fibrous=f,scaly=y,silky=k,smooth=s
* stalk-surface-below-ring: fibrous=f,scaly=y,silky=k,smooth=s
* stalk-color-above-ring: brown=n,buff=b,cinnamon=c,gray=g,orange=o, pink=p,red=e,white=w,yellow=y
* stalk-color-below-ring: brown=n,buff=b,cinnamon=c,gray=g,orange=o, pink=p,red=e,white=w,yellow=y
* veil-type: partial=p,universal=u
* veil-color: brown=n,orange=o,white=w,yellow=y
* ring-number: none=n,one=o,two=t
* ring-type: cobwebby=c,evanescent=e,flaring=f,large=l, none=n,pendant=p,sheathing=s,zone=z
* spore-print-color: black=k,brown=n,buff=b,chocolate=h,green=r, orange=o,purple=u,white=w,yellow=y
* population: abundant=a,clustered=c,numerous=n, scattered=s,several=v,solitary=y
* habitat: grasses=g,leaves=l,meadows=m,paths=p, urban=u,waste=w,woods=d
* Missing Attribute Values: 2480 of them (denoted by "?"), all for attribute #11. Class Distribution: - * edible: 4208 (51.8%) - poisonous: 3916 (48.2%) - total: 8124 instances

### Analysis Details.......
* Part 1 : Import data, clean, perofrom exploratory analysis and test the best model fit.
* Part 2 : Take a deep look into important variables and further classify what feature of the variable helps in identifiying whether the mushoom is edible or not.

### Environment Setup
```{r,include=TRUE,warning=FALSE,message=FALSE}

# Load the required packages (if packages are not available, install them first)
for (package in c('caret','readr','ggplot2','magrittr','ggthemes','dplyr','corrplot','caTools')) {
  if (!require(package, character.only=T, quietly=T)) {
    install.packages(package)
    library(package,character.only=T)
  }
}

# We will be using H2o package 
# Load H2o library into R environment
library(h2o)
# Make a connection to the h2o server
h2o.init(nthreads = -1, #Number of threads -1 means use all cores on your machine
         max_mem_size = "8G")  #max mem size is the maximum memory to allocate to H2O
h2o.init(ip="localhost", port = 54321, startH2O = TRUE)

```

### Import and Read the data - using Sys.time to keep an eye on the data parsing time
```{r,include=TRUE,warning=FALSE,message=FALSE}

start <- Sys.time()
mushrooms_csv <- "https://archive.ics.uci.edu/ml/machine-learning-databases/mushroom/agaricus-lepiota.data"
mushrooms.hex <- h2o.importFile(path = mushrooms_csv,destination_frame = "mushrooms_data.hex")
parseTime <- Sys.time() - start
print(paste("Took",round(parseTime, digits = 2),"seconds to parse", nrow(mushrooms.hex), "rows and", ncol(mushrooms.hex),"columns."))
head(mushrooms.hex)

# Since H2o stores the column names as C1,C2,C3,....... we will replace the labels to its original colnames.
names(mushrooms.hex) <- c("class","cap.shape","cap.surface","cap.color","bruises", "odor","gill.attachment","gill.spacing","gill.size","gill.color","stalk.shape","stalk.root","stalk.surface.above.ring","stalk.surface.below.ring","stalk.color.above.ring","stalk.color.below.ring", "veil.type","veil.color","ring.number","ring.type","spore.print.color","population", "habitat")
head(mushrooms.hex)

# Now Check for the Mushroom data we imported into the h2o server at http://localhost:54321/flow/index.html



```

### Exploratorory Analysis
```{r,include=TRUE,warning=FALSE,message=FALSE}

# Check for the dimensions of the data
dim(mushrooms.hex)

# Study the Structure of the Data
# 
str(mushrooms.hex)

# From the structure we can see  all the variables have Factor objects with differnet levels.
# Factors are the r-objects which are created using a vector. It stores the vector along with the distinct values of the elements in the vector as labels. The labels are always character irrespective of whether it is numeric or character or Boolean etc. in the input vector.


## Let us see how many levels each variable have 

h2o.levels(mushrooms.hex)

#h2o.unique(mushrooms.hex)
#h2o.ddply(mushrooms_hex, 2, function(x) length(unique(x)))

# Check for missing values NA's
# 
any(is.na(mushrooms.hex))

# [1] 0 - this means no NA's found 

#  1. None of the data are missing the dataset is Structured
#  2. you dont have to deal with omitting rows or columns incase there are most missing values.
#  3. you have accurate and not any predictied or average value replacing the missing data.
#  4. Less time consumption.

```

### BUILDING MODELS
```{r,include=TRUE,warning=FALSE,message=FALSE}
#This function will do the test,train and validation data split and build Random forest,GLM,GBM and Deep Learning Model.
# First, we will create three splits for train/test/valid independent data sets.
# We will train a data set on one set and use the others to test the validity
# The second set will be used for validation most of the time. The third set will
#  be withheld until the end, to ensure that our validation accuracy is consistent
#  with data we have never seen during the iterative process. 


# splits <- function(data){
splits <- h2o.splitFrame(
  mushrooms.hex,         ##  splitting the H2O frame we read above
  ratios = c(0.6,0.2),   ##  create splits of 60% and 20%; 
  #  H2O will create one more split of 1-(sum of these parameters)
  #  so we will get 0.6 / 0.2 / 1 - (0.6+0.2) = 0.6/0.2/0.2
  seed=1)                ##  setting a seed will ensure reproducible results (not R's seed)
train <- h2o.assign(splits[[1]], "train.hex")   
# assign the first result the R variable train
# and the H2O name train.hex
valid <- h2o.assign(splits[[2]], "valid.hex")   ## R valid, H2O valid.hex
test <- h2o.assign(splits[[3]], "test.hex")     ## R test, H2O test.hex

x_train = train[,2:23]
y_train = train[,1]

x_test = test[,2:23]
y_test = test[,1]

print(paste("Training data has", ncol(train),"columns and", nrow(train), "rows, whereas test data has", nrow(test), "rows, and validation data has rows", nrow(valid))
)

# Take a look at the first few rows of the data set
train[1:5,]   ## rows 1-5, all columns


# Assign X and Y values

myY <- "class"
myX <- setdiff(names(train), myY)

## Run our first predictive model (Random Forest Model)
mush_rf_model  <- h2o.randomForest(x = myX,
                 y = myY,
                 training_frame = train,
                 validation_frame = test,
                 model_id = "mush_rf_model",
                 ntrees = 250,
                 max_depth = 30,
                 seed = 100)
print(mush_rf_model)
#Let us see what variables are important in this model.
h2o.varimp_plot(mush_rf_model, num_of_features = NULL)
h2o.confusionMatrix(mush_rf_model,train)

#Fit to test and see how good it is at predicting classification!
h2o.predict(mush_rf_model,test)
h2o.confusionMatrix(mush_rf_model,test)



# Run GBM

mush_gbm_model <- h2o.gbm(x=myX,build_tree_one_node = T,
            y = myY,
            training_frame = train,
            validation_frame = test,
            model_id = "mush_gbm_model",
            ntrees = 500,
            max_depth = 6,
            learn_rate = 0.1)


# Print model performance using train data
print(mush_gbm_model)
h2o.varimp_plot(mush_gbm_model, num_of_features = NULL)

## Run DeepLearning
mush_dl_model <- h2o.deeplearning(x = myX,
                        y = myY,
                        training_frame = train,
                        validation_frame = test,
                        activation = "TanhWithDropout",
                        input_dropout_ratio = 0.2,
                        hidden_dropout_ratios = c(0.5,0.5,0.5),
                        hidden = c(50,50,50),
                        epochs = 100,
                        seed = 123456)
print(mush_dl_model)
plot(mush_dl_model)

## Performance on validation set
h2o.confusionMatrix(mush_dl_model)

# Interesting thing to note
#  Warning message:
# In .h2o.startModelJob(algo, params, h2oRestApiVersion) :
# Dropping bad and constant columns: [veil.type].
```
# Using Random Forest turns out to have more accurate classification than using tree and using bagging.
## Odor, spore.print.colo, gill.color should be important features to indicate poisonous mushroom. 
## In general, Random Forest has the best classification accuracy. Odor, spore.print.colo and stalk.color.below.ring 

### PART 2 OF THE MUSHROOM DATASET ANALYSIS 
### Since we know the top 10 features that help us to classify the Mushrooms as "Edible" or "Poisonous"
### We will a insight look at the type of sub features that clearly helps us to Classify our identification process.  

```{r,include=TRUE,warning=FALSE,message=FALSE}
# # Subset the data using the top 10 features determined by our MODEL's
# Testing feature sub classification using first 100 rows
mushrooms.hex1 <- mushrooms.hex[1:10, c(1,6,9,10,12,13,20,21,22,23)]
mushrooms.hex1
names(mushrooms.hex1) <- c("class","odor","gill.size","gill.color","stalk.root","stalk-surface-above-ring","ring.type","spore.print.color","population","habitat")
head(mushrooms.hex)
mushrooms.hex1 <- as.data.frame(mushrooms.hex1)
mushroom.hex1 <- table(mushrooms.hex1)
barplot(mushroom.hex1)

# Data Transformation
# Transforms the Class column
class_trans <- function(key){
  switch (key,
         'p' = 'poisonous',
         'e' = 'edible'
  )
}

#Transforms the odor column
odor_trans <- function(key)(
switch(key,
        'a' = 'almond',
        'l' = 'anise',
        'c'= 'creosote',
        'y'= 'fishy',
        'f'= 'foul',
        'm'= 'musty',
        'n'= 'none',
        'p'= 'pungent',
        's'= 'spicy'
   )
  )


# Transforms the gill.size column
gill.size_trans <- function(key){
  switch(key,
         'b'= 'broad',
         'n'= 'narrow')}


#Transforms the gill.color column
gill.color_trans <- function(key){
  switch(key,
         'k'= 'black',
         'n'= 'brown',
         'b'= 'buff',
         'h'= 'chocolate',
         'g'= 'gray')}


#Transforms the stalk.root column
stalk.root_trans <- function(key){
    switch(key,
           'b'= 'bulbous',
           'c'= 'club',
           'u'= 'cup',
           'e'= 'equal',
           'z'= 'rhizomorphs',
           'r'= 'rooted',
           '?'= 'missing')}

#Transforms the stalk.surface.above.ring column
stalk.surface.above.ring_trans <- function(key){
    switch(key,
           'f'= 'fibrous',
           'y'= 'scaly',
           'k'= 'silky',
           's'= 'smooth')}


# Transforms the ring.type column
ring.type_trans <- function(key){
  switch(key,
         'c' = 'cobwebby',
         'e' = 'evanescent',
         'f' = 'flaring',
         'l' = 'large',
         'n' = 'none',
         'p' = 'pendant',
         's' = 'sheathing',
         'z' = 'zone')}

#Transforms the spore.print.color column
spore.print.color_trans <- function(key){
  switch(key,
         'k'=  'black',
         'n'=  'brown',
         'b' = 'buff',
         'h'=  'chocolate',
         'r' = 'green',
         'o' = 'orange',
         'u' = 'purple',
         'w' = 'white',
         'y' = 'yellow')}

#Transforms the population column
population_trans <- function(key){
  switch(key, 
         'a' = 'abundant',
         'c' = 'clustered',
         'n' = 'numerous',
         's' = 'scattered',
         'v' = 'several',
         'y' = 'solitary')}

# Transforms the habitat column
habitat_trans <- function(key){
    switch(key,
            'g' = 'grasses',
            'l' = 'leaves',
            'm' = 'meadows',
            'p' = 'paths',
            'u' = 'urban',
            'w' = 'waster',
            'd' = 'woods')}


# Applying data transformation on the mushroom dataset

mushrooms.hex1$class <- sapply(mushrooms.hex1$class,class_trans)
mushrooms.hex1$`spore.print.color` <- sapply(mushrooms.hex1$`spore.print.color`,spore.print.color_trans)
mushrooms.hex1$`gill.color` <- sapply(mushrooms.hex1$`gill.color`,gill.color_trans)
mushrooms.hex1$`stalk.surface.above.ring` <- sapply(mushrooms.hex1$`stalk.surface.above.ring`,stalk.surface.above.ring_trans)
mushrooms.hex1$`gill.size` <- sapply(mushrooms.hex1$`gill.size`,gill.color_trans) 
mushrooms.hex1$`stalk.root` <- sapply(mushrooms.hex1$`stalk.root`,stalk.root_trans)
mushrooms.hex1$`ring.type` <- sapply(mushrooms.hex1$`ring.type`, ring.type_trans)
mushrooms.hex1$`odor` <- sapply(mushrooms.hex1$`odor`,odor_trans)
mushrooms.hex1$`population` <- sapply(mushrooms.hex1$`population`,population_trans)
mushrooms.hex1$`habitat` <- sapply(mushrooms.hex1$`habitat`,habitat_trans)

head(mushrooms.hex1)

mushroom_features <-  lapply(seq(from=2, to=ncol(mushrooms.hex1)), 
                         function(x) {table(mushrooms.hex1$class, mushrooms.hex1[,x])})
names(mushroom_features) <- colnames(mushrooms.hex1)[2:ncol(mushrooms.hex1)]
for(i in 1:length(mushroom_features)) {
  print("Deep Look at the Features")
  print(names(mushroom_features)[i])
  print(mushroom_features[[i]]) 
}

# Shut down the H2o data frame
h2o.shutdown(prompt=FALSE)

```



